Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-04-25 LICS2021-final-for-arxiv-HAL.tex
almost final

%%%%% DOMINIC: THIS IS ONLY USED FOR THE TWIN COLLAPSE EXAMPLE; PACKAGE CAN BE KILLED
%\usepackage{adjustbox}%% CONFLICT %% easier to align boxes (such as proofs) to top with, e.g. \begin{adjustbox}{valign=t,raise=-.6ex} <STUFF> \end{adjustbox} 
\usepackage{prftree}% ftp://ftp.dante.de/tex-archive/macros/latex/contrib/prftree/prftreedoc.pdf http://get-software.net/macros/latex/contrib/prftree/prftree.sty
\prflinethickness=.25pt% package default 0.2pt
\prfinterspace=1.5em% increase default space between proof rule hypotheses
\prflinepadbefore=0.2ex% default 0.3ex  above proof lines
\prflinepadafter=0ex%  default 0.3ex  below proof lines
\prflineextra=0pt% default 0.3em  proof line overhang each side

@@

\author{\IEEEauthorblockN{Dominic J. D. Hughes}
\IEEEauthorblockA{}
\and
\IEEEauthorblockN{Lutz Straßburger}
\IEEEauthorblockA{Inria, Equipe Partout\\
Ecole Polytechnique, LIX \\ %UMR 7161\\
France}
\and
\IEEEauthorblockN{Jui-Hsuan Wu}
\IEEEauthorblockA{Ecole Normale Supérieure\\
France}}

@@

\IEEEoverridecommandlockouts
\IEEEpubid{\makebox[\columnwidth]{978-1-6654-4895-6/21/\$31.00~
\copyright2021 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
  We uncover a close relationship between combinatorial and
  syntactic proofs for first-order logic. Whereas syntactic proofs are
  formalized in a deductive proof system based on inference rules,
  a combinatorial proof is a syntax-free presentation of a proof
  that is independent from any set of inference rules. We show that
  the two proof representations are related via a deep inference
  decomposition theorem that establishes a new kind of normal form
  for syntactic proofs. This yields (a) a simple proof
  of soundness and completeness for first-order combinatorial
  proofs, and (b) a full completeness theorem: every combinatorial
  proof is the image of a syntactic proof.
\end{abstract}

@@

%\todo{Examples,examples,examples}
\section{Introduction}

@@

% two syntactic proofs of 
% $\protect\peirceformula$,
% in the sequent calculus (above)
% and in the calculus of structures (below),
% which translate to the same propositional combinatorial proof (centre).
% %
% Right:
% two syntactic proofs of 
% $\protect\veedrinkerformula$,
% in the sequent calculus (above)
% and the new calculus $\FOKS$ introduced in this paper (below),
% which translate to the same first-order combinatorial proof (centre).%

@@

This is where \emph{combinatorial proofs} come in. They were
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a syntax-free notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in inductively-generated proofs
and consider the proof as a combinatorial object, more precisely
as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula=\peirceformula$ is shown mid-left in
Fig.\,\ref{fig:translate},
%depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%
% \begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
%   \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
%   \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
% \end{tikzpicture}\vspace{0ex}\end{center}
% %
% 
% \begin{equation*}
%   \scalebox{.7}{$\lkdrinkerproof
% %\qquad
% \ksonedrinkerproof
% $}
% \end{equation*}
% 
% \juihsuan{LK1 and KS1 proofs of Peirce's law and LK1 proof of drinker formula}
% \lutz{can you also do the KS1 for the drinker?}
% \juihsuan{done}

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula=\veedrinkerformula$ is shown on the
right of Fig.\,\ref{fig:translate}, %\todo{wrong implies symbol in drinkerformula?}
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} nor
\emph{full}~\cite{abramsky:jagadeesan:94}. %\dominic{ok to say sequentializable too?}
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

@@

% \subsection{Pictures that could be used elsewhere in the paper}
% 
% 
% Condensed combinatorial proof of drinker formula(s):
% %
% \begin{center}\vspace{3ex}\begin{math}
% \begin{tikzpicture}
% \drinkerInline
% \end{tikzpicture}
% \hspace{10ex}
% \begin{tikzpicture}
% \veedrinkerInline
% \end{tikzpicture}\end{math}\end{center}

%%\figcps Fig.\,\ref{fig:cps} is a floating figure with four combinatorial proofs.

%%\figcpscondensed Fig.\,\ref{fig:cps-condensed} is a floating figure with the condensed forms of the four combinatorial proofs in Fig.\,\ref{fig:cps}.

% Illustrating why we need to collapse twins during contraction, to preserve the target as a fograph:
% %
% \twincollapsepic
% %
% Aligning both $\singleton{}\mkern4mu\singleton[red]{}$ and
% $\singleton{}\mkern4mu\singleton[blue]{}$ over
% a single copy of $\forall y\mkern2mu\py$
% results in two uncoloured vertices
% \raisebox{2pt}[0ex][0ex]{$\column{\singleton{}\\[-1.6ex]\singleton{}}$\hspace{-2pt}}
% over $\forall y$.
% %
% The cover therefore fails to be a fograph: both uncoloured vertices are implicitly labelled with $y$ and
% so are outer $y$-binders in the scope of each other.
% %
% The correct operation is shown above-right,
% in which the troublesome pair is collapsed to a single uncoloured vertex $\singleton{}$ over $\forall y$.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.
%\dominic{I changed fm$(\Gamma)$ to $\form\Gamma$,
%easier for referee to understand/intuit/remember later in the paper}

@@

%The \bfit{complement} of a graph $\gG=\tuple{\vG,\eG}$ is
%the graph $\compl\gG=\tuple{\vG,\compl\eG}$ where $vw\in\compl\eG$ iff
%$vw\notin\eG$.

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\labelof v\in L$ associated with it, its \bfit{label}.  A
graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class is a
\bfit{colour}.  A \bfit{vertex renaming} of $\gG=\tuple{\vG,\eG}$
along a bijection $(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

%% Modules in cographs correspond precisely to the subtrees of the
%% cotrees (the term-trees constructing the graph via $+$ and~$\times$).\todo{Cotree
%% must alternate $+$ and $\times$. Skip this comment completely?}

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph in which all binders are legal. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}
%
\begin{remark}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. px) \vlor (\forall x. qx)$, then
$\graphof A = \single x \ \single px \ \single x \ \single qx$,
the scope of each $x$-binder contains all the vertices, in particular,
the other $x$-binder. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. px) \vlor (\exists x. qx)$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}
%
%% For this reason, we call a formula
%% \bfit{clean} if it does not contain subformulas of the form $(\forall
%% x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$,
%% \todo{need the four cases as in clear definition in fopws? Consider
%% $(\forall x.A)\vlor(B\vlor (\forall x.C))$. But then Lemma 6 doesn't hold.
%% (fopws proves only that there's a surjection from clear formulas onto fographs,
%% whose kernel is the six formula equalities)} and no
%% $x$-quantified formula occurs as subformula of another $x$-quantified
%% formula. Then we have:
%
%% \begin{lemma}
%%   If $A$ is clean iff\/ $\graphof A$ is a fograph.
%% \end{lemma}
%
%% \begin{proof}
%%   Induction on $A$, using  Theorem~\ref{thm:cograph}.
%% \end{proof}
%
%% Even though for every formula $A$ we can obtain an
%% equivalent clean formula by simply renaming some bound variables, this
%% is not unique up to $\alpha$-conversion, as is the case for
%% rectified formulas.
%
\noindent We define a congruence relation $\fequ$ on formulas by the following equations:
\begin{equation}
  \label{eq:fequ}
  \begin{array}{c@{\hskip1.5em}c}
    A \vlan B \fequ B \vlan A
    &
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) 
    \\
    A \vlor B \fequ B \vlor A
    &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C)
    \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A
    &
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B
    \\
    \exists x . \exists y . A \fequ \exists y . \exists x . A
    &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation}
where $x$ must not be free in $B$ in the last two equations.
Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$. We have the
following theorem:
\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

@@

\begin{definition}
A \bfit{first-order net} or \bfit{fonet} is a linked fograph which
has a dualizer but no induced bimatching.  
\end{definition}
%
\begin{figure}[!t]
  \begin{center}
    \vspace{-1ex}
    \begin{tikzpicture}[graph]
      \begin{scope}[xshift=-2.2cm]\twolinkfograph\end{scope}
        \begin{scope}[xshift=2.2cm]\twolinkleapgraph\end{scope}
    \end{tikzpicture}\vspace{-4ex}
  \end{center}%
  \caption{\label{fig:leap}A fonet (left) with
    dualizer $\protect\twolinkassignment$
    and its
    leap graph (right).}
\end{figure}%
%
%\noindent
Figure~\ref{fig:leap} shows a fonet with a unique dualizer, and its leap graph.

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
\liftingdiagrams
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$}, is a variable renaming
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
% is \bfit{existential-preserving} 
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is existential in $\gH$.
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labelling in the upper graph, as below-right.
  %% , which refer to as 
  %% \bfit{condensed form}.\dominic{I said ``can be inferred'' to account for 
  %% e.g.\
  %% $\graphof{\forall x_1\forall x_2 (px_i\vee px_j\vee px_k\vee px_l)}
  %% \to \graphof{\forall x\mkern2mu px}$ whose
  %% unlabelled skeleton is the skeleton of many distinct formula variants (``who binds who?'').}
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%%\subsection{First-order combinatorial proofs}
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are in
Figure~\ref{fig:cps-condensed} in \emph{condensed form}, with the formula
graph left implicit.
%  (No information is lost, since 
% atom/quantifier-occurrences in a formula
% are in bijection with vertices of its graph.)
\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\rsubstof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}

@@

% We show the equivalence of $\FOMLL$ and $\FOMLS$.
% 
% We show that $\FOMLL$ is equivalent to $\FOMLS$.
We show that $\FOMLL$ and $\FOMLS$ are equivalent.

@@

%% \begin{thm}
%%   Let $\vlder{\FOMLS}{\Deri}{A}{\ttt}$ be given. Then
%%   the graph $\graphof{\rectif A}$ forms a fonet whose linking is
%%   $\linkingof{\rectif\Deri}$ and whose dualizer is
%%   $\substof{\rectif\Deri}$.
%% \end{thm}

%% \begin{proof}
%%   Apply Lemma~\ref{lem:MLS1->fonet} to the rectification $\rectif\Deri$ of $\Deri$.
%% \end{proof}

@@

\clearpage

\appendix

>>>>

%\usepackage{adjustbox}%% CONFLICT %% easier to align boxes (such as proofs) to top with, e.g. \begin{adjustbox}{valign=t,raise=-.6ex} <STUFF> \end{adjustbox} 
\usepackage{prftree}% ftp://ftp.dante.de/tex-archive/macros/latex/contrib/prftree/prftreedoc.pdf http://get-software.net/macros/latex/contrib/prftree/prftree.sty
\prflinethickness=.25pt% package default 0.2pt
\prfinterspace=1.5em% increase default space between proof rule hypotheses
\prflinepadbefore=0.2ex% default 0.3ex  above proof lines
\prflinepadafter=0ex%  default 0.3ex  below proof lines
\prflineextra=0pt% default 0.3em  proof line overhang each side

@@

\author{
  \IEEEauthorblockN{Dominic J.\ D.\ Hughes}
  \IEEEauthorblockA{%
	Logic Group\\
	U.C.\ Berkeley\\
	USA}
\and
\IEEEauthorblockN{Lutz Straßburger}
\IEEEauthorblockA{Inria, Equipe Partout\\
Ecole Polytechnique, LIX \\ %UMR 7161\\
France}
\and
\IEEEauthorblockN{Jui-Hsuan Wu}
\IEEEauthorblockA{Ecole Normale Supérieure\\
France}}

@@

%% \newcommand{\longer}[2]{#1}
%% \newcommand{\shorter}[2]{#2}

%% %\let\choice=\longer
%% \let\choice=\shorter

\IEEEoverridecommandlockouts
\IEEEpubid{\makebox[\columnwidth]{\emph{[Long version of the LICS 2021 paper, with proofs in the appendix.]}\hfill}
    \hspace{\columnsep}\makebox[\columnwidth]{ }}
%\IEEEpubid{\makebox[\columnwidth]{978-1-6654-4895-6/21/\$31.00~
%\copyright2021 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
  We uncover a close relationship between combinatorial and
  syntactic proofs for first-order logic (without equality). Whereas syntactic proofs are
  formalized in a deductive proof system based on inference rules,
  a combinatorial proof is a syntax-free presentation of a proof
  that is independent from any set of inference rules. We show that
  the two proof representations are related via a deep inference
  decomposition theorem that establishes a new kind of normal form
  for syntactic proofs. This yields (a) a simple proof
  of soundness and completeness for first-order combinatorial
  proofs, and (b) a full completeness theorem: every combinatorial
  proof is the image of a syntactic proof.
\end{abstract}

@@

\section{Introduction}

@@

@@

This is where \emph{combinatorial proofs} come in. They were
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a syntax-free notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in inductively-generated proofs
and consider the proof as a combinatorial object, more precisely
as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula=\peirceformula$ is shown mid-left in
Fig.\,\ref{fig:translate},
%depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula=\veedrinkerformula$ is shown on the
right of Fig.\,\ref{fig:translate}, 
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} nor
\emph{full}~\cite{abramsky:jagadeesan:94}. 
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

@@

\medskip
\noindent
    {\bf Outline:} This paper has three parts. First,
    in Sections~\ref{sec:fologic}--\ref{sec:foks} we present the
    preliminaries on first-order logic, first-order graphs,
    first-order combinatorial proofs, and the first-order proof system
    $\FOKS$. Second, in Section~\ref{sec:main} we state the main
    results. And third, in
    Sections~\ref{sec:LK1-KS1}--\ref{sec:summary} we give their
    proofs.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.

@@

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\labelof v\in L$ associated with it, its \bfit{label}.
A graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class
is a \bfit{colour}.\footnote{In~\cite{hughes:fopws} it is also
required that adjacent vertices have different colours, because it
is part of the standard definition of colouring in graph
theory.
However, this condition is redundant (and therefore ommitted here), since it is implied
by the preclusion of bimatchings in the next section.}
%Therefore we do not have this additional condition here.}
A \bfit{vertex renaming} of
$\gG=\tuple{\vG,\eG}$ along a bijection
$(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph in which all binders are legal. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:

\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
\end{proof}
%
\begin{remark}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. px) \vlor (\forall x. qx)$, then
$\graphof A = \single x \ \single px \ \single x \ \single qx$,
the scope of each $x$-binder contains all the vertices, in particular,
the other $x$-binder. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. px) \vlor (\exists x. qx)$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}
%
\noindent We define a congruence relation $\fequ$ on formulas by the following equations:
\begin{equation}
  \label{eq:fequ}
  \begin{array}{c@{\hskip1.5em}c}
    A \vlan B \fequ B \vlan A
    &
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) 
    \\
    A \vlor B \fequ B \vlor A
    &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C)
    \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A
    &
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B
    \\
    \exists x . \exists y . A \fequ \exists y . \exists x . A
    &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation}
where $x$ must not be free in $B$ in the last two equations.
Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$. We have the
following theorem:
\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

@@

\begin{definition}
A \bfit{first-order net} or \bfit{fonet} is a linked fograph which
has a dualizer but no induced bimatching.  
\end{definition}
%
\begin{figure}[!t]
  \begin{center}
    \vspace{-1ex}
    \begin{tikzpicture}[graph]
      \begin{scope}[xshift=-2.2cm]\twolinkfograph\end{scope}
        \begin{scope}[xshift=2.2cm]\twolinkleapgraph\end{scope}
    \end{tikzpicture}\vspace{-4ex}
  \end{center}%
  \caption{\label{fig:leap}A fonet (left) with
    dualizer $\protect\twolinkassignment$
    and its
    leap graph (right).}
\end{figure}%
%
%\noindent
Figure~\ref{fig:leap} shows a fonet with its dualizer and leap graph.

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
%
\liftingdiagrams
%
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$}, is a variable renaming
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
%
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
%
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is existential in $\gH$.
%
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
\end{definition}
%
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labelling in the upper graph, as below-right.
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%
%
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
%
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are in
Figure~\ref{fig:cps-condensed} in \emph{condensed form}, with the formula
graph left implicit.

\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
%
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\rsubstof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
\end{remark}

@@

We show that $\FOMLL$ and $\FOMLS$ are equivalent.

@@

@@

%\clearpage
\newpage
\appendix
